NumPy---Array Programming in Python

### 2020-02-04 body.tex
DOC: Add reference for Spyder
DOC: Additional changes
Small rewrites to simplify language

\emph{Strides} are necessary to interpret computer memory, which stores elements
linearly, as multidimensional arrays.
It describes the number of bytes to move forward in memory to jump from row to
row or column to column.
Consider, for example, a 2-D array of floating-point numbers with shape
$(4, 3)$, where each element occupies 8 bytes in memory.
To move between consecutive columns we need to jump forward 8 bytes in memory,
and to access the next row $3 \times 8 = 24$ bytes.
The strides of that array are therefore $(24, 8)$.  NumPy is able to
store arrays in either C or Fortran memory order, iterating
first over either rows or columns.  This allows external libraries
written in those languages to access NumPy array data in memory directly.

@@

To complement the array syntax, NumPy includes functions that perform
\emph{vectorized} calculations on arrays, including arithmetic, statistics, and
trigonometry (Fig.~\ref{fig:array-concepts}d).
Vectorization---operating on whole arrays rather than their individual
elements---is essential to array programming.
This means that operations that would take many tens of lines to express in
languages such as C can often be implemented as a single, clear Python
expression.
This results in concise code and frees users to focus on the details of
their analysis, while NumPy handles looping over array elements near-optimally,
taking into consideration, for example, strides, in order to best utilize the
computer's fast cache memory.

@@

The interactive environment created by this array programming
foundation---along with the surrounding ecosystem of tools---inside of
IPython or Jupyter, is ideally suited to exploratory data analysis.
Users fluidly inspect, manipulate, and visualize their data, and
rapidly iterate to refine programming statements. These statements are
then stitched together into imperative or functional programs or
notebooks containing both computation and narrative.
Scientific computing beyond exploratory work is often done in a text editor
or an integrated development environment (IDEs) such as Spyder.
This rich and productive environment has made Python popular
for scientific research.

@@

% maybe mention that there are typically utilities
% to convert to and from numpy arrays
% and why (presumably to get access to the ecosystem)
The community's efforts to fill this gap led to a
proliferation of new array implementations. Each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX \cite{jax2018github} arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, and in a delayed
execution mode to allow for performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition, there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by xarray \cite{hoyer2017xarray}.

Such libraries often mimic the NumPy API, because it lowers the
barrier to entry for newcomers and provides the wider community with a
stable array programming interface. This, in turn, prevents disruptive
schisms like the divergence of Numeric and NumArray.
But exploring new ways of working with arrays is experimental by nature
and, in fact, several promising libraries---such as Theano and Caffe---have
already ceased development. And each time that a user
decides to try new technology, they have to
change import statements and ensure that the new library implements
all the parts of the NumPy API they currently use.

@@

Thanks to these developments, users are now capable, for example, of scaling
their computation to distributed systems with Dask. The protocols
\textit{compose} well, so it is even possible to take NumPy code and scale it
to multi-GPU systems with minimal code changes through Dask arrays containing
multiple CuPy arrays \cite{entschev2019}.

@@

These user-developers frequently had to write code from scratch to solve
their or their colleagues' problems---often in low-level languages
that preceded Python, like Fortran \cite{dongarra2008netlib} and C.
To them, the advantages of an interactive, high-level array library
were evident. The design of this new tool was informed by their
experiences with powerful interactive programming languages for
scientific computing such as APL \cite{iverson1962programming} and
Yorick \cite{munro1995using}, as well as commercial languages and
environments like IDL and Matlab.

@@

Even though NumPy is not part of Python's standard library,
it benefits from a good relationship with the Python developers.
Over the years, the Python language has added new features and
special syntax so that NumPy would have more succinct and 
easy to read array notation.
Since it is not part of the standard library, NumPy is able to
dictate its own release policies and development patterns.

>>>>

\emph{Strides} are necessary to interpret computer memory, which stores elements
linearly, as multidimensional arrays.
It describes the number of bytes to move forward in memory to jump from row to
row or column to column.
Consider, for example, a 2-D array of floating-point numbers with shape
$(4, 3)$, where each element occupies 8 bytes in memory.
To move between consecutive columns, we need to jump forward 8 bytes in memory,
and to access the next row $3 \times 8 = 24$ bytes.
The strides of that array are therefore $(24, 8)$.  NumPy can
store arrays in either C or Fortran memory order, iterating
first over either rows or columns.  This allows external libraries
written in those languages to access NumPy array data in memory directly.

@@

To complement the array syntax, NumPy includes functions that perform
\emph{vectorized} calculations on arrays, including arithmetic, statistics, and
trigonometry (Fig.~\ref{fig:array-concepts}d).
Vectorization---operating on whole arrays rather than their individual
elements---is essential to array programming.
This means that operations that would take many tens of lines to express in
languages such as C can often be implemented as a single, clear Python
expression.
This results in concise code and frees users to focus on the details of
their analysis, while NumPy handles looping over array elements near-optimally,
taking into consideration, for example, strides, to best utilize the
computer's fast cache memory.

@@

The interactive environment created by this array programming
foundation---along with the surrounding ecosystem of tools---inside of
IPython or Jupyter, is ideally suited to exploratory data analysis.
Users fluidly inspect, manipulate, and visualize their data, and
rapidly iterate to refine programming statements. These statements are
then stitched together into imperative or functional programs or
notebooks containing both computation and narrative.
Scientific computing beyond exploratory work is often done in a text editor
or an integrated development environment (IDEs) such as Spyder\cite{spyder2020}.
This rich and productive environment has made Python popular
for scientific research.

@@

% maybe mention that there are typically utilities
% to convert to and from numpy arrays
% and why (presumably to get access to the ecosystem)
The community's efforts to fill this gap led to a
proliferation of new array implementations. Each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX \cite{jax2018github} arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, utilizing lazy evaluation
to allow for additional performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition, there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by xarray \cite{hoyer2017xarray}.

Such libraries often mimic the NumPy API, because it lowers the
barrier to entry for newcomers and provides the wider community with a
stable array programming interface. This, in turn, prevents disruptive
schisms like the divergence of Numeric and NumArray.
But exploring new ways of working with arrays is experimental by nature
and, in fact, several promising libraries---such as Theano and Caffe---have
already ceased development. And each time that a user
decides to try new technology, they must
change import statements and ensure that the new library implements
all the parts of the NumPy API they currently use.

@@

Thanks to these developments, users can now, for example, scale
their computation from a single machine to distributed systems using Dask. The protocols
\textit{compose} well, and so it is possible to redeploy NumPy code onto multi-GPU systems with minimal changes\cite{entschev2019}.

@@

These user-developers frequently had to write code from scratch to solve
their or their colleagues' problems---often in low-level languages
that preceded Python, like Fortran \cite{dongarra2008netlib} and C.
To them, the advantages of an interactive, high-level array library
were evident. The design of this new tool was informed by their
experiences with powerful interactive programming languages for
scientific computing such as APL \cite{iverson1962programming} and
Yorick \cite{munro1995using}, as well as commercial languages and
environments like IDL and {MATLAB}.

@@

Even though NumPy is not part of Python's standard library,
it benefits from a good relationship with the Python developers.
Over the years, the Python language has added new features and
special syntax so that NumPy would have a more succinct and 
easier to read array notation\cite{smith2014pep465}.
Since it is not part of the standard library, NumPy is able to
dictate its own release policies and development patterns.
